Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-05-31 article/main.tex
re-intro

\todo{mentioner stream ciphers}

Not all pseudo-random generators are of cryptographic strength. In some
applications, it is not necessarily necessary: to be used in Monte-Carlo
simulations or generate random choices in games, a relaxed, non-cryptographic
notion of pseudo-randomness may be sufficient. This allows for faster
algorithms. For instance, the \texttt{Math.Random} function in Google's V8
open-source JavaScript engine uses the \textsf{XorShift128} generator, which is
a Linear Feedback Shift Register tailored for high software speed. On 64-bit
processors, it produces 64 pseudo-random bits with 3 shifts and 4 XORs
instructions. The \textsf{python} standard library's \texttt{random} module uses
the Mersenne Twister. The \textsf{C} library that comes along \texttt{gcc} (the
\texttt{glibc}) uses a (poor) truncated linear congruential generator by default
to implement the \texttt{rand} function.

@@

In any case, The goal of designers then consists in designing the simplest and
fastest possible pseudo-random generator that passes the day's favourite test
suite. A few selected ones are shown in Fig.~\ref{fig:prngs}. Deterministic
pseudo-random generators can be traced back to the work of Von Neumann

Lehmer proposed
linear congruential generators in 1949 for use on the ENIAC computer~\cite{Lehmer}

\texttt{Xorshift}~\cite{Marsaglia03},
\texttt{Xorshift128+}~\cite{Vigna17}.

@@

Failures in cryptographic pseudo-random generators have catastrophic security
implications. Let us mention for instance the well-known problem in Debian
Linux, where a bug in the OpenSSL package led to insufficient entropy gathering
and to practical attacks on the SSH and SSL protocols (the only remaining source
of entropy comes from the PID of the process, i.e. 16 bits or less of effective
entropy)~\cite{YilekRSES09}.

However, problems in non-cryptographic random number generators can also have
dire consequences (barring the obvious case where they are used in lieu of their
cryptographic counterparts). When they are used in scientific computing for
Monte-Carlo simulations, their defects have the potential to alter the results
of the numerical simulation. Ferrenberg et al.~\cite{Ferrenberg92} ran a
classical Ferromagnetism Ising model Monte-Carlo simulation, in a special case
where the exact results could be computed analytically, and compared the results
of the simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the physical
simulation. Different generators produced different biases: in particular a
given LFSRs yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite (in many, repeated, trials).

@@

In most cases, it is fairly easy to see that a specific non-cryptographic PRG
does not meet the cryptographic notion of pseudo-randomness, and there are few
exceptions. Most are fairly easy to predict, meaning that after having observed
a prefix of the output, it is easy to produce the next ``pseudo-random''
bits.

In this paper, we study the \textsf{PCG} family of non-cryptographic
pseudo-random generators~\cite{melissapaper,melissaweb}. Its designer claimed
that predicting it ought to be ``challenging'', even though cryptographic
strength is not claimed.

\textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
consists in applying a non-linear filtering function on top of a (fairly weak)
linear congruential generator (in a way reminiscent to the venerable filtered
LFSRs). The resulting combination is fast and passes current test suites. Its
designer claimed that distinguishing the pseudo-random stream produced would be
``challenging''. The \textsf{PCG} family contains many members, but we focus on
its strongest member, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has
a 128-bit internal state and produces 64 bits when clocked. It is the default
pseudo-random number generator in the popular \textsf{NumPy} scientific
computing package for \textsf{Python}.

@@

While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
primitive, obtaining an actual prediction algorithm requires the use of (fairly
conventional) cryptanalytic techniques. Making it practical requires in addition
a non-trivial implementation effort.

Our algorithm reconstruct the internal state using the ``guess-and-determine''
technique: some bits of the internal state are guessed ; assuming the guesses
are correct, some other information is computed ; a consistency check discards
bad guesses early on ; then candidate internal states are computed and fully
tested. The problem actually come in two distinct flavors.

@@

\paragraph{Related Work.} Knuth considered the problem of recovering the
internal state of a truncated linear congruential generator
$x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$ are unknown~\cite{Knuth85}; he
gave an algorithm exponential in the number of truncated bits.

@@

\todo{mentionner Vigna ?}

% The combination between being highly commended (it achieved 5th place out of 29
% in a recent performance test survey\cite{survey}) and a well presented
% website\cite{melissaweb} giving an feeling of professionalism can lead to the
% impression that PCG is a sufficiently secure family of generators.

% Some attacks against PCG have been carried out by Sebastiano
% Vigna\cite{vignaweb}, but this didn't yield any significant results, given that
% he only tried to break an incomplete version in a fairly straightforward way. He
% worked on a very specific version of PCG. He considers the result without
% modulo, which is a big simplification. Moreover, he didn't study the 128 bits
% long generator, but only the 64 bits long\cite{vignacode}.

@@

We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=13$. In
fact, we empirically observed that it returns the right result with probability
$\approx 66\%$ when $\ell=11$ and with probability $99\%$ when
$\ell=12$. Decrementing $\ell$ reduces the workload by a factor of 4. Therefore,
if $T$ denotes the running time when $\ell=13$, we actually used the following
strategy:
\begin{enumerate}
\item Run with $\ell=11$ (if it succeeds, stop).
\item Run with $\ell=11$ with different inputs.
\item Run with $\ell=11$ with different inputs.
\item Run with $\ell=12$.
\item Run with $\ell=13$.
\end{enumerate}
It is not difficult to check that the expected running time of this procedure is
$\frac{1079}{10800} T \approx 0.1 T$. In our implementation, $T = 200,000$ CPU
hours, so the expected running time of the reconstruction procedure is about
$20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
attempt with $\ell=11$ succeeded and took only $12,500$ CPU hours.

The whole process actually ran in 35 wall-clock minutes using 512 cluster nodes,
each equipped with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}
(``\emph{Cascade Lake}''). The actual machine is the
\verb|[----REDACTED FOR PEER REVIEW----]| located at the
\verb|[----ALSO REDACTED----]| computation center. Note that on this particular
parallel computer, running the algorithm with $\ell=13$ would take 10 hours
using the same amount of resources, so the whole procedure \emph{is} practical,
even in the absolute worst case.

@@

With $\ell=11$, the whole process took $2^{56.74}$ CPU cycles, which makes less
than 6 cycles per iteration of the inner loop. We used essentially the same
implementation tricks discussed above. However, this time we had to additionally
implement the \textsc{ConsistencyCheck} procedure, which is called in the inner
loop. We observed that the set of possible candidate values $\mathcal{C}$ only
depends on $w_0$ (the variable of the outer loop). Therefore, before entering
the inner loop, we precompute a bit field of size $2^\ell$ describing
$\mathcal{C}_i$. To simplify the implementation, we flatten them by computing
$\mathcal{C} = \cup_i \mathcal{C}_i$. This slightly increase the probability of
false positives.

>>>>

It is well-known that pseudo-random generators can be turned into symmetric
encryption algorithm, by generating ``random'' masks to be used in the one-time
pad. This is precisely what stream ciphers do.

Not all pseudo-random generators are of cryptographic strength. In some
applications, it is not necessarily necessary: to be used in Monte-Carlo
numerical simulations or generate random choices in games, a relaxed,
non-cryptographic notion of pseudo-randomness may be sufficient. This allows for
faster algorithms. For instance, \textsf{python} standard library's
\texttt{random} module uses the Mersenne Twister~\cite{MatsumotoN98}. The
\textsf{C} library that comes along \texttt{gcc} (the \texttt{glibc}) uses a
(poor) truncated linear congruential generator by default to implement the
\texttt{rand} function.

@@

In any case, designers of conventional pseudo-random generators try to obtain
the simplest and fastest algorithm that passes the day's favourite test suite. A
few selected ones are shown in Fig.~\ref{fig:prngs}. \texttt{Lehmer64} is a
truncated linear congruential generator, touted as ``the fastest PRNG that
passes \texttt{BigCrush}''~\cite{LemireLehmer}. \texttt{Xorshift128} is a clever
implementation of a 128-bit LFSR with period $2^{128} - 1$ due to
Marsaglia~\cite{Marsaglia03}, using only a few simple 32-bit
operations. \texttt{Xorshift128+} is a improved version due to
Vigna~\cite{Vigna17} that returns the sum of two consecutive outputs of a
Xorshift LFSR; it passes the ``BigCrush'' test suite and is the default PRNG in
many Javascript implementations, including that in Google's V8 engine (Chrome),
Firefox and Safari.

@@

Failures in cryptographic pseudo-random generators have catastrophic security
implications. Let us mention for instance the well-known problem in Debian
Linux from 2008, where a bug in the OpenSSL package led to insufficient
entropy gathering and to practical attacks on the SSH and SSL protocols (the
only remaining source of entropy comes from the PID of the process,
\textit{i.e.} 16 bits or less of effective entropy)~\cite{YilekRSES09}.

However, problems in non-cryptographic random number generators can also have
dire consequences (barring the obvious case where they are used in lieu of their
cryptographic counterparts). When they are used in scientific computing for
Monte-Carlo methods, their defects have the potential to alter the results of
numerical simulations. Ferrenberg et al.~\cite{Ferrenberg92} ran a classical
Ferromagnetism Ising model Monte-Carlo simulation, in a special case where the
exact results could be computed analytically, and compared the results of the
simulation with the ``true'' answer. They used several conventional
pseudo-random generators: a 32-bit linear congruential generator, two LFSRs,
various combinations thereof, etc. They observed that changing the source of
random numbers significantly altered the outcome of the numerical
simulation. Different generators produced different biases: in particular a
given LFSRs yielded energy levels that were systematically too low and critical
temperatures that were always too high, while another kind of generator yielded
the opposite (in many, repeated, trials).

@@

In most cases, it is fairly easy to see that a specific non-cryptographic PRG
does not meet the cryptographic notion of pseudo-randomness, and there are few
exceptions. Most are fairly easy to predict, meaning that after having observed
a prefix of the output, it is easy to produce the next ``pseudo-random''
bits. This makes a good source of exercises for cryptography students.

In this paper, we study the \textsf{PCG} family of non-cryptographic
pseudo-random generators proposed by O'Neil~\cite{melissapaper,melissaweb}. She
did not claim that the algorithm has cryptographic strength, but that predicting
its output ought to be ``challenging''. We therefore took up the challenge.

\textsf{PCG} stands for ``Permuted Congruential Generator'': it essentially
consists in applying a non-linear filtering function on top of a linear
congruential generator (in a way reminiscent to the venerable filtered
LFSRs). The resulting combination is fast and passes current statistical test
suites. The \textsf{PCG} family contains many members, but we focus on the
strongest one, named either \textsf{PCG64} or \textsf{PCG-XSL-RR}. It has a
128-bit internal state and produces 64 bits when clocked. It is the default
pseudo-random number generator in the popular \textsf{NumPy} scientific
computing package for \textsf{Python}.

@@

While the \textsf{PCG} pseudo-random generator is not meant as a cryptographic
primitive, obtaining an actual prediction algorithm requires the use of
cryptanalytic techniques. Making it practical requires in addition a non-trivial
implementation effort.

Our algorithm reconstruct the internal state using a ``guess-and-determine''
approach: some bits of the internal state are guessed ; assuming the guesses are
correct, some other information is computed ; a consistency check discards bad
guesses early on ; then candidate internal states are computed and fully
tested. The problem actually come in two distinct flavors.

@@

\paragraph{Related Work.} Deterministic pseudo-random generators can be traced
back to the work of Von Neumann and Metropolis on the ENIAC
computer~\cite{vonNeumann1951}; they suggested around 1946 to use the
``middle-square'' method: if $u_n$ is a $k$-digit number, form $u_{n+1}$ by
taking the square of the $\frac{k}{2}$ middle digits of $u_n$. This is a
venerable precursor of the Blum-Blum-Shub ``provably secure'' PRNG. The main
problem of this method is that it produces sequences that quickly enters short
cycles.

Lehmer later proposed linear congruential generators in 1949, also for use on
the ENIAC computer~\cite{Lehmer}. He gave the sequence defined by
$u_0 = 47594118, u_{n+1} = 23 u_n \bmod 10^8 + 1$ and proved that it had period
5882353, a clear improvement compared to the middle-square approach. More
details on early pseudo-random generators can be found in~\cite{Knuth}.

Knuth considered the problem of recovering the internal state of a truncated
linear congruential generator $x_{i+1} = ax_i + c \bmod 2^k$ when $a$ and $c$
are unknown~\cite{Knuth85}; he gave an algorithm exponential in the number of
truncated bits.

@@

@@

We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=13$. In
fact, we empirically observed that it returns the right result with probability
$\approx 66\%$ when $\ell=11$ and with probability $99\%$ when
$\ell=12$. Decrementing $\ell$ reduces the workload by a factor of 4. Therefore,
if $T$ denotes the running time when $\ell=13$, we actually used the following
strategy:
\begin{algorithmic}[1]
\State Run with $\ell=11$ (if it succeeds, stop).
\State Run with $\ell=11$ with different inputs.
\State Run with $\ell=11$ with different inputs.
\State Run with $\ell=12$.
\State Run with $\ell=13$.
\end{algorithmic}
It is not difficult to check that the expected running time of this procedure is
$\frac{1079}{10800} T \approx 0.1 T$. In our implementation, $T = 200,000$ CPU
hours, so the expected running time of the reconstruction procedure is about
$20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
attempt with $\ell=11$ succeeded and took only $12,500$ CPU hours.

The whole process actually ran in 35 wall-clock minutes using 512 cluster nodes,
each equipped with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}
(``\emph{Cascade Lake}''). The actual machine is the
\verb|[----REDACTED FOR PEER REVIEW----]| computer located at the
\verb|[----ALSO REDACTED----]| computation center. Note that on this particular
parallel computer, running the algorithm with $\ell=13$ would take 10 hours
using the same amount of resources, so the whole procedure \emph{is} practical,
even in the absolute worst case.

@@

With $\ell=11$, the whole process took $2^{56.74}$ CPU cycles, which makes less
than 6 cycles per iteration of the inner loop. We used essentially the same
implementation tricks discussed above. However, this time we had to additionally
implement the \textsc{ConsistencyCheck} procedure, which is called in the inner
loop. We observed that the set of possible candidate values $\mathcal{C}$ only
depends on $w_0$ (the variable of the outer loop). Therefore, before entering
the inner loop, we precompute a bit field of size $2^\ell$ describing
$\mathcal{C}_i$. To simplify the implementation, we flatten them by computing
$\mathcal{C} = \cup_i \mathcal{C}_i$. This slightly increase the probability of
false positives, but makes our code slightly simpler.
